---
title: "report-COVID-19"
author: "Thomas Gredig"
date: "3/17/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message  = FALSE)
library(ggplot2)
library(reshape2)
library(knitr)
scientific_10 <- function(x) {
    xout <- gsub("1e", "10^{", format(x),fixed=TRUE)
    xout <- gsub("{-0", "{-", xout,fixed=TRUE)
    xout <- gsub("{+", "{", xout,fixed=TRUE)
    xout <- gsub("{0", "{", xout,fixed=TRUE)
    xout <- paste(xout,"}",sep="")
    return(parse(text=xout))
}
scale_y_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_y_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
scale_x_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_x_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
```

# COVID-19

We are comparing the growth rates in different regions. Using a 1-fit exponential model, the confirmed cases generally fit the model well. The doubling time is then compared.

The data is from the [CSSE COVID-19 Dataset](https://github.com/CSSEGISandData/COVID-19) after analysis of [Coronavirus by Tomas Pueyo](https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca). 


## Time Series

Regions with more than 100 confirmed cases increased rapidly from Mar 1 to Mar 11:

```{r}
path.series = 'csse_covid_19_data/csse_covid_19_time_series'
file.list = dir(path.series, pattern='csv$')
filename = file.list[1]
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')

NUM.CASES = 300
SEL.DATE = '2020-03-01'
m2 = subset(m1, Date==SEL.DATE)
m3 = aggregate(m2$Confirmed, by=list(m2$Region), FUN=sum)
names(m3) = c('Region','Confirmed')
m3 = subset(m3, Confirmed>NUM.CASES)
ggplot(m3, aes(x=reorder(Region, Confirmed), y=Confirmed)) + 
  geom_bar(stat='identity', fill='red') +
  xlab('') +
  ggtitle(paste("More than",NUM.CASES,"confirmed COVID-19 cases:",SEL.DATE),subtitle = 'Source: Johns Hopkins CSSE') +
  scale_y_log10nice() + 
  coord_flip() + 
  theme_bw() 
```

Only **some days later** the list of countries hugely expanded, so on `r max(m1$Date)`, we have:

```{r}
SEL.DATE = max(m1$Date)
m2 = subset(m1, Date==SEL.DATE)
m3 = aggregate(m2$Confirmed, by=list(m2$Region), FUN=sum)
names(m3) = c('Region','Confirmed')
m3 = subset(m3, Confirmed>NUM.CASES)
#m3$Region = as.character(levels(m3$Region)[m3$Region])
ggplot(m3, aes(x=reorder(Region, Confirmed), y=Confirmed)) + 
  geom_bar(stat='identity', fill='red') + 
  xlab('') +
  ggtitle(paste("More than",NUM.CASES,"confirmed COVID-19 cases:",SEL.DATE),subtitle = 'Source: Johns Hopkins CSSE') + 
  scale_y_log10nice() + 
  coord_flip() + 
  theme_bw() 
top.regions = levels(m3$Region)[m3$Region]
```


```{r}
m4 = aggregate(m1$Confirmed, by=list(m1$Region, m1$Date), FUN=sum)
names(m4) = c('Region','Date','Confirmed')
m4p = which(m4$Region %in% top.regions)
m5 = m4[m4p,]
ggplot(m5, aes(Date, Confirmed, col=Region)) + 
  geom_line() + 
  scale_y_log10nice(limits=c(100,2e5),name='confirmed COVID-19 cases') + ylab('') + 
  theme_bw()
```


Select a few countries with exponential growth in comparison with China:

```{r}
ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | 
                Region=='Switzerland' | Region=='Iran' | Region=='China' | Region=='Spain'), aes(Date, Confirmed, col=Region)) + 
  geom_point() + 
  geom_smooth(se=FALSE) + #, method="lm", formula= (Confirmed ~ exp(Date))) + 
  scale_y_log10nice(limits=c(100,2e5),name='confirmed COVID-19 cases') + 
  theme_bw()
ggsave('confirmed-COVID-19-selectedRegions.png', width=6,height=4, dpi=150)
```


Renormalize the growth starting with the 100th confirmed case. Here is the list of countries:

```{r}
m5a = subset(m5, Confirmed >= 100)
m5a$DateDays = round(as.numeric(difftime( m5a$Date, '2020-02-14', units='days')))
q.dates = aggregate(m5a$Date, by=list(m5a$Region), FUN=min)
names(q.dates) = c('Region','Date')
kable(q.dates[order(q.dates$Date),])
```

Make a graph with the trajectories from that point onwards:

```{r}
m8 = data.frame()
for(q in 1:nrow(q.dates)) {
  if (!(q.dates$Region[q] %in% c('Cruise Ship','Denmark','Qatar'))) {
    d1 = subset(m5a, Region == q.dates$Region[q])
    d1$DateDaysNorm = round(as.numeric(d1$DateDays - difftime(q.dates$Date[q],'2020-02-14', units='days')))
    m8 = rbind(m8, d1)
  }
}
ggplot(m8, aes(DateDaysNorm, Confirmed, col=Region)) + 
  geom_line(size=1, alpha=0.5) + 
  geom_point(data=subset(m8,Region=='US'), size=3) +
  scale_y_log10nice(name='Confirmed COVID-19 Cases') + 
  scale_x_continuous(limits=c(0,22)) + xlab('Days since 100th confirmed case') + 
  theme_bw()
ggsave('renormalized-COVID-19-confirmed.png',width=6,height=4, dpi=150)
```

Add an exponential fit:

```{r}
m8a = subset(m8, Region == 'US')
A0 = m8a$Confirmed[1]
nls(data = m8a,
    Confirmed ~ A0*exp(DateDaysNorm/T1),
    start = list( T1=3)) -> fit
m8b = data.frame(DateDaysNorm = 0:20,
                 Confirmed = predict(fit, list(DateDaysNorm=0:20)))
T1 = signif(summary(fit)$coeff[1],2)
ggplot(m8a, aes(DateDaysNorm, Confirmed)) + 
  geom_point(col='red', size=2) + 
  geom_line(data=m8b) + ylab('Confirmed COVID-19 cases') + 
  ggtitle(paste('US since',min(m8a$Date),' doubling time:',T1,'days')) + xlab('days') + 
  scale_y_continuous(limits=c(0,1e4)) + 
  scale_x_continuous(limits=c(0,20), breaks=0:4*7, minor_breaks=0:30) +
  theme_bw()
summary(fit)
```

Semi-log plot:

```{r}
ggplot(m8a, aes(DateDaysNorm, Confirmed)) + 
  geom_point(col='red', size=2) + 
  geom_line(data=m8b) + ylab('confirmed COVID-19 cases') + 
  ggtitle(paste('US since',min(m8a$Date),' doubling time:',T1,'days')) + xlab('days') +   scale_y_log10nice(name='Confirmed COVID-19 cases') + 
  scale_x_continuous(limits=c(0,20), breaks=0:4*7, minor_breaks=0:30) +
  theme_bw()
```


## Growth rates

Growth rates since 100th confirmed case in different countries for first 10 days.

```{r}
result = data.frame()
for (r in levels(droplevels(m8$Region))) {
  print(r)
  m8a = subset(m8, Region == r)
  m8a = subset(m8a, DateDaysNorm <= 10)
  if (nrow(m8a)>5) {
    A0 = m8a$Confirmed[1]
    nls(data = m8a,
      Confirmed ~ A0*exp(DateDaysNorm/T1),
      start = list(T1=3)) -> fit
    r1 =  data.frame(
      Region = r,
      T = summary(fit)$coeff[1],
      T.sd = summary(fit)$coeff[2],
      len = nrow(m8a)
      )
    result=rbind(result,r1)
  }
}
ggplot(result, aes(x=reorder(Region,-T), T)) + 
  #geom_bar(stat='identity', fill='red') +
  geom_point(col='red', size=3) + 
  geom_errorbar(aes(ymax=T+T.sd, ymin=T-T.sd)) + xlab('Region') +
  scale_y_continuous(limits=c(0,14)) + ylab('confirmed cases doubling time (days)') + 
  coord_flip()
```





## Deaths

Graphing some countries with exponentially growing death rates at the moment:

```{r}
filename = file.list[2]
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')
m4 = aggregate(m1$Confirmed, by=list(m1$Region, m1$Date), FUN=sum)
names(m4) = c('Region','Date','Confirmed')
m4p = which(m4$Region %in% top.regions)
m5 = m4[m4p,]
#ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | 
#                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(Date, Confirmed, #col=Region, shape=Region)) + 
#  geom_point() + 
#  scale_y_log10nice(limits=c(2,2e4)) + 
#  scale_shape_manual(values=1:nlevels(m5$Region)) +
#  theme_bw()
```




```{r}
m5$DateDays = round(as.numeric(difftime( m5$Date, '2020-02-14', units='days')))
ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | Region=='China' |
                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(DateDays, Confirmed, col=Region, shape=Region)) + 
  geom_point() + 
  scale_x_continuous(limits=c(-30,35)) + xlab('days since Valentine')+ 
  scale_y_log10nice(limits=c(2,2e3), name='COVID-19 Deaths') + ylab('Deaths') + 
  scale_shape_manual(values=1:nlevels(m5$Region)) +
  theme_bw()
```


Renormalizing the start dates for `recorded deaths`, we find the dates when the `5th death` was recorded in each country:

```{r}
m6= subset(m5, Confirmed>5)
q1 = aggregate(m6$DateDays, by=list(m6$Region), FUN=min)
names(q1) = c('Country','Start')
kable(q1[order(q1$Start),])
```


Renormalizing the start dates for `recorded deaths`, we find the dates when the `15th death` was recorded in each country:

```{r}
m6= subset(m5, Confirmed>15)
q1 = aggregate(m6$DateDays, by=list(m6$Region), FUN=min)
names(q1) = c('Country','Start')
kable(q1[order(q1$Start),])
```



```{r}
m7 = data.frame()
for(q in 1:nrow(q1)) {
  d1 = subset(m6, Region == q1$Country[q])
  d1$DateDaysNorm = d1$DateDays - q1$Start[q]
  m7 = rbind(m7, d1)
}
ggplot(subset(m7, Region=='US' | Region=='France' | Region=='Italy' | Region=='China' |
                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(DateDaysNorm, Confirmed, col=Region)) + 
  geom_point() + geom_path() + 
  scale_x_continuous(limits=c(-0,30)) + xlab('days since 15th death in region')+ 
  scale_y_log10nice(limits=c(12,3e3), name='COVID-19 Deaths') + ylab('Deaths') + 
  scale_shape_manual(values=1:nlevels(m5$Region)) +
  theme_bw()
ggsave('renormalized-COVID-19-deaths.png',width=5,height=3,dpi=150)
```