---
title: "report-COVID-19"
author: "Thomas Gredig"
date: "3/12/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message  = FALSE)
library(ggplot2)
library(reshape2)
library(knitr)
scientific_10 <- function(x) {
    xout <- gsub("1e", "10^{", format(x),fixed=TRUE)
    xout <- gsub("{-0", "{-", xout,fixed=TRUE)
    xout <- gsub("{+", "{", xout,fixed=TRUE)
    xout <- gsub("{0", "{", xout,fixed=TRUE)
    xout <- paste(xout,"}",sep="")
    return(parse(text=xout))
}
scale_y_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_y_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
scale_x_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_x_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
```

# COVID-19

We determine the order of countries it spread using the exponential growth model and determine that it started in China, then spread to Iran, South Korea, Italy, U.S., France, and then Spain based on confirmed deaths.

Japan and South Korea have a much slower growth in the death rate (see graph at the bottom). China, Italy, and U.S. appear to have common trajectories.

The data is from the [CSSE COVID-19 Dataset](https://github.com/CSSEGISandData/COVID-19) after analysis of [Coronavirus by Tomas Pueyo](https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca). 


## Time Series

Regions with more than 100 confirmed cases increased rapidly from Mar 1 to Mar 11:

```{r}
path.series = 'csse_covid_19_data/csse_covid_19_time_series'
file.list = dir(path.series, pattern='csv$')
filename = file.list[1]
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')

NUM.CASES = 300
SEL.DATE = '2020-03-01'
m2 = subset(m1, Date==SEL.DATE)
m3 = aggregate(m2$Confirmed, by=list(m2$Region), FUN=sum)
names(m3) = c('Region','Confirmed')
m3 = subset(m3, Confirmed>NUM.CASES)
ggplot(m3, aes(x=reorder(Region, Confirmed), y=Confirmed)) + 
  geom_bar(stat='identity', fill='red') +
  xlab('') +
  ggtitle(paste("More than",NUM.CASES,"confirmed COVID-19 cases:",SEL.DATE),subtitle = 'Source: Johns Hopkins CSSE') +
  scale_y_log10nice() + 
  coord_flip() + 
  theme_bw() 
```

Only **some days later** the list of countries hugely expanded, so on `r max(m1$Date)`, we have:

```{r}
SEL.DATE = max(m1$Date)
m2 = subset(m1, Date==SEL.DATE)
m3 = aggregate(m2$Confirmed, by=list(m2$Region), FUN=sum)
names(m3) = c('Region','Confirmed')
m3 = subset(m3, Confirmed>NUM.CASES)
#m3$Region = as.character(levels(m3$Region)[m3$Region])
ggplot(m3, aes(x=reorder(Region, Confirmed), y=Confirmed)) + 
  geom_bar(stat='identity', fill='red') + 
  xlab('') +
  ggtitle(paste("More than",NUM.CASES,"confirmed COVID-19 cases:",SEL.DATE),subtitle = 'Source: Johns Hopkins CSSE') + 
  scale_y_log10nice() + 
  coord_flip() + 
  theme_bw() 
top.regions = levels(m3$Region)[m3$Region]
```


```{r}
m4 = aggregate(m1$Confirmed, by=list(m1$Region, m1$Date), FUN=sum)
names(m4) = c('Region','Date','Confirmed')
m4p = which(m4$Region %in% top.regions)
m5 = m4[m4p,]
ggplot(m5, aes(Date, Confirmed, col=Region)) + 
  geom_point() + 
  scale_y_log10nice(limits=c(100,2e5),name='confirmed COVID-19 cases') + ylab('') + 
  theme_bw()
```


Select a few countries with exponential growth in comparison with China:

```{r}
ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | 
                Region=='Switzerland' | Region=='Iran' | Region=='China' | Region=='Spain'), aes(Date, Confirmed, col=Region)) + 
  geom_point() + 
  geom_smooth(se=FALSE) + #, method="lm", formula= (Confirmed ~ exp(Date))) + 
  scale_y_log10nice(limits=c(100,2e5),name='confirmed COVID-19 cases') + 
  theme_bw()
ggsave('confirmed-COVID-19-selectedRegions.png', width=6,height=4, dpi=150)
```


Renormalize the growth starting with the 100th confirmed case. Here is the list of countries:

```{r}
m5a = subset(m5, Confirmed >= 100)
m5a$DateDays = round(as.numeric(difftime( m5a$Date, '2020-02-14', units='days')))
q.dates = aggregate(m5a$Date, by=list(m5a$Region), FUN=min)
names(q.dates) = c('Region','Date')
kable(q.dates[order(q.dates$Date),])
```

Make a graph with the trajectories from that point onwards:

```{r}
m8 = data.frame()
for(q in 1:nrow(q.dates)) {
  if (!(q.dates$Region[q] %in% c('Cruise Ship','Denmark','Qatar'))) {
    d1 = subset(m5a, Region == q.dates$Region[q])
    d1$DateDaysNorm = round(as.numeric(d1$DateDays - difftime(q.dates$Date[q],'2020-02-14', units='days')))
    m8 = rbind(m8, d1)
  }
}
ggplot(m8, aes(DateDaysNorm, Confirmed, col=Region)) + 
  geom_line(size=1, alpha=0.5) + 
  geom_point(data=subset(m8,Region=='US'), size=3) +
  scale_y_log10nice(name='Confirmed COVID-19 Cases') + 
  scale_x_continuous(limits=c(0,22)) + xlab('Days since 100th confirmed case') + 
  theme_bw()
ggsave('renormalized-COVID-19-confirmed.png',width=6,height=4, dpi=150)
```

Add an exponential fit:

```{r}
m8a = subset(m8, Region == 'US')
nls(data = m8a,
    Confirmed ~ A + B*exp(DateDaysNorm/T1),
    start = list(A=-100, B=100, T1=3)) -> fit
m8b = data.frame(DateDaysNorm = 0:20,
                 Confirmed = predict(fit, list(DateDaysNorm=0:20)))
ggplot(m8a, aes(DateDaysNorm, Confirmed)) + 
  geom_point(col='red') + 
  geom_line(data=m8b) + ylab('Confirmed COVID-19 cases') + 
  ggtitle(paste('US since',min(m8a$Date))) + xlab('days') + 
  scale_y_continuous(limits=c(0,1e4)) + 
  scale_x_continuous(limits=c(0,20), breaks=0:4*7, minor_breaks=0:30)
summary(fit)
```


## Growth rates

Growth rates since 100th confirmed case in different countries for first 10 days.

```{r}
result = data.frame()
for (r in levels(droplevels(m8$Region))) {
  m8a = subset(m8, Region == r)
  m8a = subset(m8a, DateDaysNorm <= 11)
  if (nrow(m8a)>5) {
    if (r=='Japan') {
      nls(data = m8a,
        Confirmed ~ -100 + B*exp(DateDaysNorm/T1),
        start = list( B=100, T1=13)) -> fit    
        r1 =  data.frame(
      Region = r,
      T = summary(fit)$coeff[2],
      T.sd = summary(fit)$coeff[4],
      ampl = summary(fit)$coeff[1],
      len = nrow(m8a)
      )
    } else {
      nls(data = m8a,
        Confirmed ~ A + B*exp(DateDaysNorm/T1),
        start = list(A=-100, B=100, T1=3)) -> fit
    r1 =  data.frame(
      Region = r,
      T = summary(fit)$coeff[3],
      T.sd = summary(fit)$coeff[6],
      ampl = summary(fit)$coeff[2],
      len = nrow(m8a)
      )
    }
    result=rbind(result,r1)
  }
}
ggplot(result, aes(x=reorder(Region,-T), T)) + 
  #geom_bar(stat='identity', fill='red') +
  geom_point(col='red', size=3) + 
  geom_errorbar(aes(ymax=T+T.sd, ymin=T-T.sd)) + xlab('Region') +
  scale_y_continuous(limits=c(0,20)) + ylab('confirmed cases doubling time (days)') + 
  coord_flip()

ggplot(result, aes(x=reorder(Region,ampl), ampl)) + 
  geom_bar(stat='identity', fill='red') +
  xlab('Region') + ylab('Amplitude Growth for Confirmed Cases in first 11 days') + 
  coord_flip()
```

```{r}
ggplot(result, aes(T, ampl, label=Region)) +
  geom_point(size=3, col='red') +
  geom_point(data=subset(result, Region=='US'),size=2,col='black') +
  geom_text(check_overlap = TRUE, vjust=-0.8) +
  xlab('confirmed cases doubling time (days)')  +
  ylab('confirmed case amplitude')
```

## Deaths

Graphing some countries with exponentially growing death rates at the moment:

```{r}
filename = file.list[2]
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')
m4 = aggregate(m1$Confirmed, by=list(m1$Region, m1$Date), FUN=sum)
names(m4) = c('Region','Date','Confirmed')
m4p = which(m4$Region %in% top.regions)
m5 = m4[m4p,]
#ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | 
#                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(Date, Confirmed, #col=Region, shape=Region)) + 
#  geom_point() + 
#  scale_y_log10nice(limits=c(2,2e4)) + 
#  scale_shape_manual(values=1:nlevels(m5$Region)) +
#  theme_bw()
```




```{r}
m5$DateDays = round(as.numeric(difftime( m5$Date, '2020-02-14', units='days')))
ggplot(subset(m5, Region=='US' | Region=='France' | Region=='Italy' | Region=='China' |
                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(DateDays, Confirmed, col=Region, shape=Region)) + 
  geom_point() + 
  scale_x_continuous(limits=c(-30,35)) + xlab('Days since Valentine')+ 
  scale_y_log10nice(limits=c(2,2e3), name='COVID-19 Deaths') + ylab('Deaths') + 
  scale_shape_manual(values=1:nlevels(m5$Region)) +
  theme_bw()
```


Renormalizing the start dates for `recorded deaths`, we find the dates when the `5th death` was recorded in each country:

```{r}
m6= subset(m5, Confirmed>5)
q1 = aggregate(m6$DateDays, by=list(m6$Region), FUN=min)
names(q1) = c('Country','Start')
kable(q1[order(q1$Start),])
```


Renormalizing the start dates for `recorded deaths`, we find the dates when the `15th death` was recorded in each country:

```{r}
m6= subset(m5, Confirmed>15)
q1 = aggregate(m6$DateDays, by=list(m6$Region), FUN=min)
names(q1) = c('Country','Start')
kable(q1[order(q1$Start),])
```



```{r}
m7 = data.frame()
for(q in 1:nrow(q1)) {
  d1 = subset(m6, Region == q1$Country[q])
  d1$DateDaysNorm = d1$DateDays - q1$Start[q]
  m7 = rbind(m7, d1)
}
ggplot(subset(m7, Region=='US' | Region=='France' | Region=='Italy' | Region=='China' |
                Region=='Switzerland' | Region=='Iran' |  Region=='Spain' | Region=='Korea, South'), aes(DateDaysNorm, Confirmed, col=Region)) + 
  geom_point() + geom_path() + 
  scale_x_continuous(limits=c(-0,30)) + xlab('Days since 15th death in region')+ 
  scale_y_log10nice(limits=c(12,3e3), name='COVID-19 Deaths') + ylab('Deaths') + 
  scale_shape_manual(values=1:nlevels(m5$Region)) +
  theme_bw()
ggsave('renormalized-COVID-19-deaths.png',width=5,height=3,dpi=150)
```