---
title: "report-COVID-19"
author: "Thomas Gredig"
date: "7/12/2020"
output: 
  rmarkdown::github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message  = FALSE)
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(reshape2)
library(knitr)
library(dplyr)
library(lubridate)
library(zoo)
scientific_10 <- function(x) {
    xout <- gsub("1e", "10^{", format(x),fixed=TRUE)
    xout <- gsub("{-0", "{-", xout,fixed=TRUE)
    xout <- gsub("{+", "{", xout,fixed=TRUE)
    xout <- gsub("{0", "{", xout,fixed=TRUE)
    xout <- paste(xout,"}",sep="")
    return(parse(text=xout))
}
scale_y_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_y_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
scale_x_log10nice <- function(name=NULL,omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_x_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}
DATA.SOURCE = '@i7ed, source: JHU CSSE COVID-19 Data'
```



# COVID-19

We are comparing the growth rates in different regions.

The data is from the [CSSE COVID-19 Dataset](https://github.com/CSSEGISandData/COVID-19).

Projections based on [SIR models](https://www.maa.org/press/periodicals/loci/joma/the-sir-model-for-spread-of-disease-the-differential-equation-model) from [HealthData](https://covid19.healthdata.org/projections).


```{r}
# return population of a country
dc1 = read.csv('data/population-by-country-2020.csv', stringsAsFactors = FALSE)
dc2 = read.csv('data/population-by-state-2020.csv', stringsAsFactors = FALSE)
dc = data.frame(
  Country = c(dc1$Country,dc2$State),
  Popoulation.2020 = c(dc1$Popoulation.2020, dc2$Population.2020),
  stringsAsFactors = FALSE
)
dc$Popoulation.2020 = as.numeric(dc$Popoulation.2020)
getPopulation <- function(countryName) {
  #countryName = gsub('\\*','',countryName)
  if (countryName=='US') { countryName='United States' }
  if (countryName=='Czechia') { countryName ='Czech Republic'}
  if (countryName=='Korea, South') { countryName = 'South Korea' }
  mc = which(dc$Country==countryName)
  if (length(mc)>= 1) { p = dc$Popoulation.2020[mc] } else { p=1e14 }
  p/1e6
}
```

```{r}
# load confirmed cases
path.series = 'csse_covid_19_data' #/csse_covid_19_time_series'
file.list = dir(path.series, pattern='csv$')
filename = 'time_series_covid19_confirmed_global.csv' 
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')
m1 %>% 
  group_by(Region, Date) %>%
  summarize(tot = sum(Confirmed)) -> m5
m1 = data.frame(
  Region = m5$Region,
  Date = as.Date(m5$Date,format='%m.%d.%y'),
  Confirmed = m5$tot
)
m1=m1[order(m1$Region, m1$Date),]
m1$Confirmed.PerMio = unlist(by(m1, m1$Region, 
                                function(x) {x$Confirmed / getPopulation(levels(x$Region)[x$Region[1]])}))
```


```{r}
# load confirmed cases in US by state
file.list = dir(path.series, pattern='csv$')
filename = 'time_series_covid19_confirmed_US.csv' 
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1:6,8:11)]
m2 = melt(d1)
names(m2) = c('Region','Date','Confirmed')
m2$Date = gsub('^X','',as.character(levels(m2$Date)[m2$Date]))
m2$Date = as.Date(m2$Date, format='%m.%d.%y')
m2 %>% 
  group_by(Region, Date) %>%
  summarize(tot = sum(Confirmed)) -> m3
m4 = data.frame(
  Region = m3$Region,
  Date = as.Date(m3$Date,format='%m.%d.%y'),
  Confirmed = m3$tot
)
m4=m4[order(m4$Region, m4$Date),]
m4$Confirmed.PerMio = unlist(by(m4, m4$Region, 
                                function(x) {x$Confirmed / getPopulation(levels(x$Region)[x$Region])}))
```


# Confirmed COVID-19 Cases 

## World

Confirmed cases over time for different regions:

```{r}
AVG.DAYS=7
m = rbind(m1, m4)
m %>%
  filter(Region == 'US' | Region=='Korea, South' | Region=='Canada' | Region == 'Germany' | Region == 'Switzerland' | Region =='Sweden') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-02')) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new cases per million') +
  geom_line(size=1.5) +
  #scale_y_log10() + 
  scale_x_date() +
  ggtitle('Confirmed new COVID-19 Cases per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


Some of the larger countries in the world:

```{r}
AVG.DAYS=7
m = rbind(m1, m4)
m %>%
  filter(Region == 'US' | Region=='Korea, South' | Region=='Canada' | Region == 'Germany' | Region == 'India' | Region =='Japan' | 
           Region == 'China' | Region == 'Nigeria' | Region == 'Russia' | Region == 'Brazil') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-02')) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new cases per million') +
  geom_line(size=1.5) +
  #scale_y_log10() + 
  scale_x_date() +
  ggtitle('Confirmed new COVID-19 Cases per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


Japan, Switzerland and U.S. on log-scale:

```{r}
AVG.DAYS=7
m = rbind(m1, m4)
m %>%
  filter(Region == 'US' | Region=='Japan' | Region == 'Switzerland' | Region == 'Korea, South') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-02')) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new cases per million') +
  geom_line(size=1.5) +
  scale_y_log10() + 
  scale_x_date() +
  ggtitle('Confirmed new COVID-19 Cases per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


## United States

Some selected regions within the U.S.


```{r}
AVG.DAYS=7
m %>%
  filter(Region == 'US' | Region=='New York' | Region=='Arizona' | Region == 'California' |  Region == 'Ohio' |
           Region == 'Florida' | Region == 'Texas') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-06')) %>%
  filter(Confirmed.3DayMean>1.1) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new cases per million') +
  geom_line(size=1.5) +
  #scale_y_continuous(breaks=c(0:5*250)) + 
  scale_x_date() +
  scale_y_log10() + 
  ggtitle('Confirmed new COVID-19 Cases per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```



```{r}
AVG.DAYS=7
m %>%
  filter(Region == 'Japan' | Region=='Florida' | Region=='Arizona' | Region == 'California' |  Region == 'France' ) %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-06')) %>%
  filter(Confirmed.3DayMean>0.1) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new cases per million') +
  geom_line(size=1.5) +
  #scale_y_continuous(breaks=c(0:5*250)) + 
  scale_x_date() +
  #scale_y_log10() + 
  ggtitle('Confirmed new COVID-19 Cases per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


# COVID-19 DEATHS



```{r}
# load deaths reported
file.list = dir(path.series, pattern='csv$')
filename = 'time_series_covid19_deaths_global.csv' 
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1,3,4)]
m1 = melt(d1)
names(m1) = c('Region','Date','Confirmed')
m1$Date = gsub('^X','',as.character(levels(m1$Date)[m1$Date]))
m1$Date = as.Date(m1$Date, format='%m.%d.%y')
m1 %>% 
  group_by(Region, Date) %>%
  summarize(tot = sum(Confirmed)) -> m5
m1 = data.frame(
  Region = m5$Region,
  Date = as.Date(m5$Date,format='%m.%d.%y'),
  Confirmed = m5$tot
)
m1=m1[order(m1$Region, m1$Date),]
m1$Confirmed.PerMio = unlist(by(m1, m1$Region, 
                                function(x) {x$Confirmed / getPopulation(levels(x$Region)[x$Region[1]])}))
```


```{r}
# load confirmed cases in US by state
file.list = dir(path.series, pattern='csv$')
filename = 'time_series_covid19_deaths_US.csv' 
f = file.path(path.series, filename)
d = read.csv(f)
d1 = d[,-c(1:6,8:11)]
m2 = melt(d1)
names(m2) = c('Region','Date','Confirmed')
m2$Date = gsub('^X','',as.character(levels(m2$Date)[m2$Date]))
m2$Date = as.Date(m2$Date, format='%m.%d.%y')
m2 %>% 
  group_by(Region, Date) %>%
  summarize(tot = sum(Confirmed)) -> m3
m4 = data.frame(
  Region = m3$Region,
  Date = as.Date(m3$Date,format='%m.%d.%y'),
  Confirmed = m3$tot
)
m4=m4[order(m4$Region, m4$Date),]
m4$Confirmed.PerMio = unlist(by(m4, m4$Region, 
                                function(x) {x$Confirmed / getPopulation(levels(x$Region)[x$Region])}))
```


## World

Statistics of some of the larger countries including Brazil:

```{r}
AVG.DAYS=7
m4$USstate = TRUE
m1$USstate = FALSE
m = rbind(m1, m4)
m %>%
  filter(Region == 'US' | Region=='Korea, South' | Region=='Canada' | 
           Region == 'Germany' | Region == 'India' | Region =='Japan' | 
           Region == 'China' | Region == 'Nigeria' | Region == 'Russia' | Region == 'Brazil') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-02')) %>%
  filter(Confirmed.3DayMean>0.01) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new deaths per million') +
  geom_line(size=1.5) +
  scale_y_log10() + 
  scale_x_date() +
  ggtitle('Confirmed new COVID-19 Deaths per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


## U.S.

Statistics of selected U.S. states

```{r}
AVG.DAYS=7
m %>%
  filter(Region == 'US' | Region=='New York' | Region=='Arizona' | Region == 'California' |  Region == 'Ohio' |
           Region == 'Florida' | Region == 'Texas') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-06')) %>%
  filter(Confirmed.3DayMean>0.1) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new deaths per million') +
  geom_line(size=1.5) +
  #scale_y_continuous(breaks=c(0:5*250)) + 
  scale_x_date() +
  scale_y_log10() + 
  ggtitle('Confirmed new COVID-19 Deaths per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```


The 3 most populous states in the U.S.

```{r}
AVG.DAYS=7
m %>%
  filter(Region == 'California' | Region == 'Florida' | Region == 'Texas' |
           Region == 'France' | Region == 'New York' | Region == 'Italy' |
           Region == 'Switzerland') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-06')) %>%
  filter(Confirmed.3DayMean>0.07) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new deaths per million') +
  geom_line(size=1.5) +
  #scale_y_continuous(breaks=c(0:5*250)) + 
  scale_x_date() +
  scale_y_log10() + 
  ggtitle('Confirmed new COVID-19 Deaths per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```

```{r}
AVG.DAYS=7
m %>%
  filter(Region == 'California' | Region == 'New York' | Region == 'Switzerland' | Region =='Germany' |
           Region == 'US') %>%
  mutate(Confirmed.PerMio.New = c(0,diff(Confirmed.PerMio))) %>%
  mutate(Confirmed.3DayMean = rollmean(Confirmed.PerMio.New, k=AVG.DAYS, fill=NA)) %>%
  filter(Date >= as.Date('2020-03-11')) %>%
  filter(Confirmed.3DayMean>0.005) %>%
  filter(Date < Sys.Date()-AVG.DAYS) %>%
  ggplot(aes(Date,Confirmed.3DayMean, color=Region)) + 
  ylab('confirmed new deaths per million') +
  geom_line(size=1.5) +
  #scale_y_continuous(breaks=c(0:5*250)) + 
  scale_x_date() +
  scale_y_log10() + 
  ggtitle('Confirmed new COVID-19 Deaths per Million', subtitle = 
            paste(DATA.SOURCE,', ',AVG.DAYS,'day average')) + 
  theme(plot.background = element_rect(fill="darkslateblue",colour='white'),
        panel.background  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text = element_text(color='white'),
        legend.background = element_blank(),
        legend.key = element_rect(fill="darkslateblue",colour='darkslateblue'),
        legend.text = element_text(colour = "white"),
        title = element_text(colour = "white"),
        axis.ticks.length.x.bottom = unit(4,"mm"),
        axis.ticks.x = element_line(color='white', linetype = 'dotted')
        )
```

## Death Rates by Country and U.S. States

As computed of `r Sys.Date()`:

```{r}
m6 = m[!is.na(m$Date),]
m %>%
  filter(is.na(Confirmed.PerMio)==FALSE) %>%
  filter(is.na(Date)==FALSE) %>%
  filter(Date == max(Date)) %>%
  filter(Confirmed.PerMio>1000) %>%
  ggplot(aes(x=reorder(Region, Confirmed.PerMio), y=Confirmed.PerMio, fill=factor(USstate))) + 
  geom_bar(stat='identity') + 
  geom_bar(data=subset(m, Region=='US' & Date==max(m6$Date)), stat='identity', fill='green') + 
  coord_flip() + xlab('Region') + ylab('Confirmed COVID-19 Deaths per Million') +
  theme(legend.position = 'none')
```

```{r}
m %>%
  filter(is.na(Confirmed.PerMio)==FALSE) %>%
  filter(is.na(Date)==FALSE) %>%
  filter(Date == max(Date)) %>%
  filter(Confirmed.PerMio>500 & Confirmed.PerMio<1000) %>%
  ggplot(aes(x=reorder(Region, Confirmed.PerMio), y=Confirmed.PerMio, fill=factor(USstate))) + 
  geom_bar(stat='identity') + 
  coord_flip() + xlab('Region') + ylab('Confirmed COVID-19 Deaths per Million') +
    theme(legend.position = 'none')
```



All countries:

```{r}
m0 = as.character(levels(m$Region)[m$Region])
m %>%
  filter(is.na(Confirmed.PerMio)==FALSE) %>%
  filter(is.na(Date)==FALSE) %>%
  filter(Date == max(Date)) %>%
  filter(Confirmed.PerMio>400) %>%
  filter(USstate==FALSE) %>%
  ggplot(aes(x=reorder(Region, Confirmed.PerMio), y=Confirmed.PerMio)) + 
  geom_bar(stat='identity') + 
  geom_bar(data=subset(m, Region=='US' & Date==max(m6$Date)), stat='identity', fill='green') +
  coord_flip() + xlab('Region') + ylab('Confirmed COVID-19 Deaths per Million') +
  theme(legend.position = 'none')
```